---
title: "Wombat-GIT-ANCOM-R"
author: "Erin D'Agnese"
date: "1/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set working directory
```{r}
setwd("C:/Users/erdag/OneDrive/Documents/Wombat")
```

Load necessary libraries
```{r}
library(exactRankTests)
library(nlme)
library(dplyr)
library(ggplot2)
library(compositions)
library(readr)
library(tidyverse)
```
Make sure the script for the ANCOM analysis ANCOM.R.2 by Frederick Huang is in your working directory or link it with code from his GitHub repository then load the script into your workspace.
```{r}
source("ancom_v2.1.R")
```

Following the Read.me file from the ANCOM.R.2 documentation use the appropriate OTU/ASV table and the corresponding Metadata file
```{r}
otu_data = read_csv("ps.colon.rarefied.genus.otu.csv")
otu_id = otu_data$`ASV.id`
otu_data = data.frame(otu_data[, -1], check.names = FALSE)
rownames(otu_data) = otu_id

meta_data = read_csv("Wombat-GIT-Metadata.csv")
meta_data = meta_data %>% rename(Sample.ID = 'sampleid')

```

create the feature table for ANCOM

```{r}
feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL 
```

#due to the number of samples per individual the use of a group to determine differentially abund taxa based on structural zeros led to identification of non significantly representative OTUs

```{r}
out_cut = 0.05; zero_cut = 0.97; lib_cut = 5000; neg_lb = TRUE
prepro = feature_table_pre_process(feature_table, meta_data, sample_var, group_var, 
                                   out_cut, zero_cut, lib_cut, neg_lb)
feature_table = prepro$feature_table # Preprocessed feature table
meta_data = prepro$meta_data # Preprocessed metadata
struc_zero = prepro$structure_zeros # Structural zero info
```

run the ancom with the main variable as Sampletype while accounting for variability at the species level

```{r}
main_var = "sampletype"; p_adj_method = "BH"; alpha = 0.05
adj_formula = "species"; rand_formula = NULL
t_start = Sys.time()
res = ANCOM(feature_table, meta_data, struc_zero, main_var, p_adj_method, 
            alpha, adj_formula, rand_formula)
t_end = Sys.time()
t_run = t_end - t_start

write_csv(res$out, "ps.colon.rarefied.sampletype.by.species.csv")

```

visualise the volcano plot of the output

```{r}
n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")

# Annotation data
dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.6"], label = "W[0.6]")

fig = res$fig +
  geom_hline(yintercept = cut_off["detected_0.6"], linetype = "dashed") +
  geom_text(data = dat_ann, aes(x = x, y = y, label = label),
            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
fig
```

run the ANCOM for species level differences and visualise

```{r}
main_var = "species"; p_adj_method = "BH"; alpha = 0.05
adj_formula = NULL; rand_formula = NULL
t_start = Sys.time()
res = ANCOM(feature_table, meta_data, struc_zero, main_var, p_adj_method, 
            alpha, adj_formula, rand_formula)
t_end = Sys.time()
t_run = t_end - t_start

write_csv(res$out, "ps.colon.rarefied.species.ancom.csv")

n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")

# Annotation data
dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.6"], label = "W[0.6]")

fig = res$fig +
  geom_hline(yintercept = cut_off["detected_0.6"], linetype = "dashed") +
  geom_text(data = dat_ann, aes(x = x, y = y, label = label),
            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
fig
```

Run the ANCOM for sampleposition and visualise

```{r}
main_var = "sample_position"; p_adj_method = "BH"; alpha = 0.05
adj_formula = "species"; rand_formula = NULL
t_start = Sys.time()
res = ANCOM(feature_table, meta_data, struc_zero, main_var, p_adj_method, 
            alpha, adj_formula, rand_formula)
t_end = Sys.time()
t_run = t_end - t_start

write_csv(res$out, "ps.colon.rarefied.sampleposition.ancom.csv")

n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")

# Annotation data
dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.6"], label = "W[0.6]")

fig = res$fig +
  geom_hline(yintercept = cut_off["detected_0.6"], linetype = "dashed") +
  geom_text(data = dat_ann, aes(x = x, y = y, label = label),
            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
fig

```

Mapping the ANCOM results onto the classification of the representative ASVs

```{r}
df1<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.rarefied.species.ancom.csv",head=T)
df2<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.rarefied.genus.mapping.csv",head=T)
species_asv_ancom<-merge(df1,df2,by=c('taxa_id'),all.x=TRUE, all.y=TRUE)
write.csv(species_asv_ancom, "C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_species_ancom_mapped.csv", row.names = FALSE)

df3<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.rarefied.sampletype.by.species.csv",head=T)
sampletype_asv_ancom<-merge(df3,df2,by=c('taxa_id'),all.x=TRUE, all.y=TRUE)
write.csv(sampletype_asv_ancom, "C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_sampletype_ancom_mapped.csv", row.names = FALSE)

df4<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.rarefied.sampleposition.ancom.csv",head=T)
sampleposition_asv_ancom<-merge(df4,df2,by=c('taxa_id'),all.x=TRUE, all.y=TRUE)
write.csv(sampleposition_asv_ancom, "C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_sampleposition_ancom_mapped.csv", row.names = FALSE)

```

Need to make sums of the reads for each rarefied ASV by variable
so need to load another package

```{r}
library(doBy)
```

Need to subset the Metadata file to only represent the colon samples

```{r}
met<-read.csv('Wombat-GIT-Metadata.csv')
meta_data = met %>% rename(Sample.ID = 'sampleid')
meta_data_HG <- subset(meta_data, meta_data$colon=='yes', select = c("Sample.ID", "species", "sampletype", "colon", "sample_position"))

```
Get metadata for the subset

```{r}
m1a<-read.csv('Wombat-GIT-Metadata.csv',nrows=1)
m1<-meta_data_HG
colnames(m1)<-colnames(m1a)
```

get the OTU data for the subset which was previously created and used in the ANCOM

```{r}
d1 <- read.csv('ps.colon.rarefied.genus.otu.csv')
d2<-d1[,-1]
OTUs<-as.character(d1[,1])
rownames(d2)<-OTUs
```

Sum reads by species

```{r}
for (j in OTUs){
  d3<-d2[j,]
  d4<-data.frame(t(d3) )
  colnames(d4)<-'OTU'
  d4$meta<-factor(m1$species)
  q3<-summaryBy(OTU ~ meta,
                data=d4,
                FUN=sum)
  if(j==OTUs[1]){q5<-q3}else{q5<-cbind(q5,q3$OTU.sum)}
}
 colnames(q5)[-1] <- as.character(OTUs)
 write.csv(q5,file='ps.colon.species.sums.csv')
```

Sum reads by sampletype

```{r}
for (j in OTUs){
  d3<-d2[j,]
  d4<-data.frame(t(d3) )
  colnames(d4)<-'OTU'
  d4$meta<-factor(m1$colon)
  q3<-summaryBy(OTU ~ meta,
                data=d4,
                FUN=sum)
  if(j==OTUs[1]){q5<-q3}else{q5<-cbind(q5,q3$OTU.sum)}
}
colnames(q5)[-1] <- as.character(OTUs)
write.csv(q5,file='ps.colon.rarefied.sampletype.sums.csv')
```
Sum reads by sampleposition
```{r}
for (j in OTUs){
  d3<-d2[j,]
  d4<-data.frame(t(d3) )
  colnames(d4)<-'OTU'
  d4$meta<-factor(m1$sample_position)
  q3<-summaryBy(OTU ~ meta,
                data=d4,
                FUN=sum)
  if(j==OTUs[1]){q5<-q3}else{q5<-cbind(q5,q3$OTU.sum)}
}
colnames(q5)[-1] <- as.character(OTUs)
write.csv(q5,file='ps.colon.rarefied.sampleposition.sums.csv')
```
Need to transpose the results and change heading for the ASV column to 'taxa_id' 
before mapping the ANCOM results with the classification mapped file to the sums files

mapping the sums onto the ANCOM results

```{r}
df5<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_species_ancom_mapped.csv",head=T)
df6<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.species.sums.csv",head=T)
HGspecies_ancom_sum<-merge(df5,df6,by=c('taxa_id'),all.x=TRUE, all.y=TRUE)
write.csv(HGspecies_ancom_sum, "C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_species_ANCOM_sums.csv", row.names = FALSE)

df7<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_sampletype_ancom_mapped.csv",head=T)
df8<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.rarefied.sampletype.sums.csv",head=T)
HGsampletype_ancom_sum<-merge(df7,df8,by=c('taxa_id'),all.x=TRUE, all.y=TRUE)
write.csv(HGsampletype_ancom_sum, "C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_sampletype_ANCOM_sums.csv", row.names = FALSE)

df9<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_sampleposition_ancom_mapped.csv",head=T)
df10<-read.csv("C:/Users/erdag/OneDrive/Documents/Wombat/ps.colon.rarefied.sampleposition.sums.csv",head=T)
HGsampleposition_ancom_sum<-merge(df9,df10,by=c('taxa_id'),all.x=TRUE, all.y=TRUE)
write.csv(HGsampleposition_ancom_sum, "C:/Users/erdag/OneDrive/Documents/Wombat/colon_rarefied_sampleposition_ANCOM_sums.csv", row.names = FALSE)

```

